---
title: "bikeMarkdown"
author: "Manny"
date: "2023-03-01"
output: html_document
---

```{r}
options(repos = list(CRAN="http://cran.rstudio.com/"))
```

```{r}
install.packages("tidyverse")
install.packages("plyr")
install.packages("dplyr")
```

```{r}
library(tidyverse)
library(plyr)
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidyr)
```

# Read in and combine all csv files from Q1 location to a dataframe
```{r}
setwd("D:/Coursera/Data Analytics/Portfolio & Case Study/Case Studies/Case Study 1_How does a bike_share navigate speedy success/Dataset/Unzipped/2020_2023/2022/Q1")
q1_2022 <- ldply(list.files(), read.csv, header=TRUE)
View(q1_2022)
```

# Read in and combine all csv files from Q2 location to a dataframe
```{r}
setwd("D:/Coursera/Data Analytics/Portfolio & Case Study/Case Studies/Case Study 1_How does a bike_share navigate speedy success/Dataset/Unzipped/2020_2023/2022/Q2")
q2_2022 <- ldply(list.files(), read.csv, header=TRUE)
View(q2_2022)
```

# Read in and combine all csv files from Q3 location to a dataframe
```{r}
setwd("D:/Coursera/Data Analytics/Portfolio & Case Study/Case Studies/Case Study 1_How does a bike_share navigate speedy success/Dataset/Unzipped/2020_2023/2022/Q3")
q3_2022 <- ldply(list.files(), read.csv, header=TRUE)
View(q3_2022)
```

# Read in and combine all csv files from Q4 location to a dataframe
```{r}
setwd("D:/Coursera/Data Analytics/Portfolio & Case Study/Case Studies/Case Study 1_How does a bike_share navigate speedy success/Dataset/Unzipped/2020_2023/2022/Q4")
q4_2022 <- ldply(list.files(), read.csv, header=TRUE)
View(q4_2022)
```

# Combine all the dataframes together and form the full year in one dataframe
```{r}
year_2022 <- bind_rows(q1_2022,q2_2022,q3_2022,q4_2022)
View(year_2022)
```

# Remove the last 4 columns (lat, long, gender, birthyear)
```{r}
year_2022 <- year_2022 %>%
  select(-c(start_lat,start_lng,end_lat,end_lng))
View(year_2022)
```
# Inspect the new table that has been created

#List of column names
```{r}
colnames(year_2022)  
```

```{r}
nrow(year_2022)  #How many rows are in data frame?
```

```{r}
dim(year_2022)  #Dimensions of the data frame?
```

```{r}
head(year_2022)  #See the first 6 rows of data frame.  Also tail(all_trips)
```

```{r}
str(year_2022)  #See list of columns and data types (numeric, character, etc)
```

```{r}
summary(year_2022)  #Statistical summary of data. Mainly for numerics
```

# Change the values
```{r}
year_2022 <-  year_2022 %>%
  mutate(member_casual = recode(member_casual
                           ,"Subscriber" = "member"
                           ,"Customer" = "casual"))
View(year_2022)
```

# check the data
```{r}
table(year_2022$member_casual)
```

# Add columns that list the date, month, day, and year of each ride
```{r}
year_2022$date <- as.Date(year_2022$started_at) #The default format is yyyy-mm-dd
```

```{r}
year_2022$month <- format(as.Date(year_2022$date), "%m")
year_2022$day <- format(as.Date(year_2022$date), "%d")
year_2022$year <- format(as.Date(year_2022$date), "%Y")
year_2022$day_of_week <- format(as.Date(year_2022$date), "%A")
```

# Add a "ride_length" calculation to all_trips (in seconds)
```{r}
year_2022$ride_length <- difftime(year_2022$ended_at,year_2022$started_at)
View(year_2022)
```

# Convert "ride_length" from Factor to numeric so we can run calculations on the data
```{r}
is.factor(year_2022$ride_length)
year_2022$ride_length <- as.numeric(as.character(year_2022$ride_length))
is.numeric(year_2022$ride_length)
```
# Remove "bad" data
# The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative
# We will create a new version of the dataframe (v2) since data is being removed
```{r}
year_2022_v2 <- year_2022[!(year_2022$start_station_name == "HQ QR" | year_2022$ride_length<0),]
View(year_2022_v2)
```
# Remove trips that are longer than a day
```{r}
year_2022 <- year_2022_v2[!(year_2022_v2$ride_length > 86400 | year_2022_v2$ride_length <= 0),]
View(year_2022)
```

# mean for year_2022_v2
```{r}
mean(year_2022_v2$ride_length) #straight average (total ride length / rides)
```

# mean for year_2022
```{r}
mean(year_2022$ride_length) #straight average (total ride length / rides)
```

# median for 2022 v2
```{r}
median(year_2022_v2$ride_length) #midpoint number in the ascending array of ride lengths
```

# median for 2022
```{r}
median(year_2022$ride_length) #midpoint number in the ascending array of ride lengths
```

# max for year 2022_v2
```{r}
max(year_2022_v2$ride_length) #longest ride
```

#longest ride for year_2022
```{r}
max(year_2022$ride_length) 
```

# min ride for year 2022 v2
```{r}
min(year_2022_v2$ride_length)
```

# min ride for year 2022 
```{r}
min(year_2022$ride_length)
```

# combine all above together year 2022 v2
```{r}
summary(year_2022_v2$ride_length)
```

# combine all above together year 2022
```{r}
summary(year_2022$ride_length)
```

# Compare members and casual users year 2022 v2
```{r}
aggregate(year_2022_v2$ride_length ~ year_2022_v2$member_casual, FUN = mean)
```

# Compare members and casual users mean year 2022
```{r}
aggregate(year_2022$ride_length ~ year_2022$member_casual, FUN = mean)
```

# Compare members and casual users median year 2022 v2
```{r}
aggregate(year_2022_v2$ride_length ~ year_2022_v2$member_casual, FUN = median)
```

# Compare members and casual users median year 2022
```{r}
aggregate(year_2022$ride_length ~ year_2022$member_casual, FUN = median)
```

# Compare members and casual users max year 2022 v2
```{r}
aggregate(year_2022_v2$ride_length ~ year_2022_v2$member_casual, FUN = max)
```

# Compare members and casual users max year 2022
```{r}
aggregate(year_2022$ride_length ~ year_2022$member_casual, FUN = max)
```

# Compare members and casual users min year 2022 v2
```{r}
aggregate(year_2022_v2$ride_length ~ year_2022_v2$member_casual, FUN = min)
```

# Compare members and casual users min year 2022
```{r}
aggregate(year_2022$ride_length ~ year_2022$member_casual, FUN = min)
```

# See the average ride time by each day for members vs casual users
```{r}
aggregate(year_2022$ride_length ~ year_2022$member_casual + year_2022$day_of_week, FUN = mean)
```

# Notice that the days of the week are out of order. Let's fix that. year 2022
```{r}
year_2022$day_of_week <- ordered(year_2022$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

# average ride time by each day for members vs casual users year 2022
```{r}
aggregate(year_2022$ride_length ~ year_2022$member_casual + year_2022$day_of_week, FUN = mean)
```

# year 2022 amount of riders average ride length
```{r}
year_2022 %>% 
  mutate(weekday = wday(started_at)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  dplyr::summarise(number_of_rides = n()							#calculates the number of rides and average duration 
  ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)								# sorts
```

# Let's visualize the number of rides by rider type year 2022
```{r}
year_2022 %>% 
  mutate(weekday = wday(started_at)) %>% 
  group_by(member_casual, weekday) %>% 
  dplyr::summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(title = "# of rides by rider type", caption = "Manuel Aguilar | Capstone Project")
```


# Let's create a visualization for average duration
```{r}
year_2022 %>% 
  mutate(weekday = wday(started_at)) %>% 
  group_by(member_casual, weekday) %>% 
  dplyr::summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(title = "Average Ride", caption = "Manuel Aguilar | Capstone Project")
```

# Create a csv file that we will visualize in Excel, Tableau, or my presentation software
```{r}
counts <- aggregate(year_2022$ride_length ~ year_2022$member_casual + year_2022$day_of_week, FUN = mean)
write.csv(counts, file = '~/bikeshare/avg_ride_length1.csv')
```

# Number of rides between members and casual riders for each month
```{r}
year_2022 %>% 
  group_by(member_casual, month) %>% 
  summarise(number_of_rides = n(), .groups = 'drop') %>% 
  arrange(month)
```

```{r}
# Comparing general bike type preference between members and casual riders
year_2022 %>% 
  group_by(rideable_type, member_casual) %>% 
  summarize(number_of_rides = n(), .groups = 'drop')
```



```{r}
my_theme = theme(plot.title=element_text(size=20),
                 axis.text.x=element_text(angle=45, size=16), 
                 axis.text.y=element_text(size=16),
                 axis.title.x=element_text(size=18), 
                 axis.title.y=element_text(size=18),
                 strip.text.x=element_text(size=16), 
                 strip.text.y=element_text(size=16),
                 legend.title=element_text(size=18), 
                 legend.text=element_text(size=16))
```


# Viz 1 - general average riding duration between members and casual riders
```{r}
options(repr.plot.width = 6, repr.plot.height = 8)

year_2022 %>% 
  group_by(member_casual) %>% 
  summarize(average_duration = mean(ride_length)) %>% 
  ggplot(aes(x = member_casual, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(x = "Rider Type", y = "Average Duration", 
       title = "Average Riding Duration by Rider Type") + my_theme
```

# Viz 2 - average riding duration of each day of week between members and casual riders
```{r}
options(repr.plot.width = 10, repr.plot.height = 8)

year_2022 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(average_duration = mean(ride_length), .groups = 'drop') %>% 
  #arrange(member_casual, day_of_week) %>% 
  ggplot(aes(x = day_of_week, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(x = "Day of Week", y = "Average Duration", 
       fill = "Member/Casual",
       title = "Average Riding Duration by Day: Members vs. Casual Riders") + my_theme
```


# Viz 3 - average number of rides of each day of week between members and casual riders
```{r}
year_2022 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n(), .groups = 'drop') %>% 
  #arrange(member_casual, day_of_week) %>% 
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + scale_y_continuous(labels = scales::comma) +
  labs(x = "Day of Week", y = "Number of Rides", fill = "Member/Casual",
       title = "Average Number of Rides by Day: Members vs. Casual Riders") + my_theme
```

# Viz 4 - average number of rides by month (casual riders)
```{r}
options(repr.plot.width = 10, repr.plot.height = 8)
year_2022 %>% 
  group_by(month, member_casual) %>% 
  summarize(number_of_rides = n(), .groups = 'drop') %>% 
  filter(member_casual == 'casual') %>%
  drop_na() %>%
  ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) + 
  geom_bar(position = 'dodge', stat = 'identity') + scale_y_continuous(labels = scales::comma) +
  theme(axis.text.x = element_text(angle = 45)) + 
  labs(x = "Month", y = "Number of Rides", 
       fill = "Member/Casual",
       title = "Average Number of Rides by Month: Casual Riders") + my_theme
```

# Viz 9 - usage of different bikes by rider type
```{r}
options(repr.plot.width = 12, repr.plot.height = 8)

year_2022 %>% 
  group_by(rideable_type, member_casual) %>% 
  summarize(number_of_rides = n(), .groups = 'drop') %>% 
  drop_na() %>% 
  ggplot(aes(x = member_casual, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(~rideable_type) +
  labs(fill = "Member/Casual", x = "", y = "Number of Rides", 
       title = "Usage of Different Bikes: Members vs. Casual Riders") + my_theme
```


```{r}
year_2022 %>%
  group_by(member_casual)%>%
  summarize(count=length(ride_id)/1000000)%>%
  mutate(pct=round(count*100/sum(count),2))%>%
  ggplot(aes(x="",y=count,fill=member_casual))+geom_bar(width=1,stat="identity",color="black",show.legend=FALSE)+coord_polar("y",start=0)+geom_text(aes(label=paste(member_casual,paste(pct,"%"),sep="\n")),position=position_stack(vjust=0.5),color="black")+labs(title="Percent of rides by rider type",caption = "Manuel Aguilar - Capstone project")+theme_void()
```

```{r}
year_2022 %>%
  group_by(member_casual)%>%
  summarize(number_of_rides = n(), .groups = 'drop')%>%
  mutate(pct=number_of_rides,2)%>%
  ggplot(aes(x="",y=number_of_rides,fill=member_casual))+geom_bar(width=1,stat="identity",color="black",show.legend=FALSE)+coord_polar("y",start=0)+geom_text(aes(label=paste(member_casual,paste(pct),sep="\n")),position=position_stack(vjust=0.5),color="black")+labs(title="Total of rides by rider type",caption = "Manuel Aguilar - Capstone project")+theme_void()
```

```{r}
options(repr.plot.width = 10, repr.plot.height = 8)
year_2022 %>% 
  group_by(month, member_casual) %>% 
  summarize(number_of_rides = n(), .groups = 'drop') %>% 
  drop_na() %>%
  ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) + 
  geom_bar(position = 'dodge', stat = 'identity') + scale_y_continuous(labels = scales::comma) +
  theme(axis.text.x = element_text(angle = 45)) + 
  labs(x = "Month", y = "Number of Rides", 
       fill = "Member/Casual",
       title = "Average Number of Rides by Month: Casual Riders") + my_theme
```

```{r}
options(repr.plot.width = 10, repr.plot.height = 8)

year_2022 %>% 
  group_by(member_casual, month) %>% 
  summarise(average_duration = mean(ride_length), .groups = 'drop') %>% 
  #arrange(member_casual, day_of_week) %>% 
  ggplot(aes(x = month, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(x = "Month", y = "Average Duration", 
       fill = "Member/Casual",
       title = "Average Riding Duration by Month: Members vs. Casual Riders") + my_theme
```

# Average duration of ride daily compared
```{r}
options(repr.plot.width = 10, repr.plot.height = 8)

year_2022 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(average_duration = mean(ride_length), .groups = 'drop') %>% 
  #arrange(member_casual, day_of_week) %>% 
  ggplot(aes(x = day_of_week, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(x = "Day of Week", y = "Average Duration", 
       fill = "Member/Casual",
       title = "Average Riding Duration by Day: Members vs. Casual Riders") + my_theme
```

